---
title: "Geospatial Data - Recreating John Snow's 1854 map of the Broad Street cholera outbreak."
format: pdf
editor: visual
---


# ------------------------------------------------------------------------------
# resources

* [Modern Data Science with R (3e) - Chapter 17](https://mdsr-book.github.io/mdsr3e/)

* [1854 Broad Street cholera outbreak](https://en.wikipedia.org/wiki/1854_Broad_Street_cholera_outbreak)

* [Shapefile](https://en.wikipedia.org/wiki/Shapefile)


# ------------------------------------------------------------------------------
# packages

```{r}
library(tidyverse)
library(sf)
library(ggspatial)
library(maps)
library(mapproj)
```


# ------------------------------------------------------------------------------
# data

```{r}

root <- here::here("/cloud/project")  # where we stored these files

dsn <- fs::path(root, "SnowGIS_SHP")

```


```{r}

list.files(dsn)

```


```{r}

st_layers(dsn)

```


```{r}

CholeraDeaths <- st_read(dsn, layer = "Cholera_Deaths")

```


```{r}

class(CholeraDeaths)

```


```{r}

plot(CholeraDeaths["Count"])

```


```{r}

ggplot(CholeraDeaths) +
  geom_sf()

```


```{r}

ggplot(CholeraDeaths) + 
  ggspatial::annotation_map_tile(type = "osm", zoomin = 0, progress = "none") + 
  geom_sf(aes(size = Count), alpha = 0.7)

```


```{r}

# bounding boxes

st_bbox(CholeraDeaths)

```


# ------------------------------------------------------------------------------
# projections

* [Top 10 World Map Projections](https://futuremaps.com/blogs/news/top-10-world-map-projections)


```{r}

maps::map(database = "world", projection = "mercator", wrap = TRUE)

```


```{r}

maps::map(database = "world", projection = "cylequalarea", param = 45, wrap = TRUE)

```


# ------------------------------------------------------------------------------
# CRS - coordinate reference system

* [Coordinate Reference System](https://en.wikipedia.org/wiki/Spatial_reference_system) - three formats that are common for storing information about the projection of a geospatial object are EPSG, PROJ.4, and WKT

* Common CRSs

  - [EPSG:4326](https://epsg.io/4326) - also known as WGS84, standard for GPS systems and Google Earth
  
  - [EPSG:3857](https://epsg.io/3857) - a Mercator projection used in map tiles by Google Maps, Open Street Maps, etc.
  
  - [EPSG:27700](https://epsg.io/27700) - also known as OSGB 1936, or the British National Grid, commonly used in Britain


```{r}
st_crs(4326)$epsg

st_crs(3857)$epsg

st_crs(27700)$epsg
```

```{r}
st_crs(4326)$Wkt

st_crs(3857)$Wkt

st_crs(27700)$Wkt
```

```{r}
st_crs(4326)$proj4string

st_crs(3857)$proj4string

st_crs(27700)$proj4string
```


#### what about our dataset?

```{r}

st_crs(CholeraDeaths)

```


```{r}

cholera_4326 <- CholeraDeaths |>
  st_transform(4326)

st_bbox(cholera_4326)

```


```{r}

ggplot(cholera_4326) + 
  annotation_map_tile(type = "osm", zoomin = 0) + 
  geom_sf(aes(size = Count), alpha = 0.7)

```


```{r}

# add pump locations

pumps <- st_read(dsn, layer = "Pumps")

pumps_4326 <- pumps |>
  st_transform(4326)

ggplot(cholera_4326) + 
  annotation_map_tile(type = "osm", zoomin = 0) + 
  geom_sf(aes(size = Count), alpha = 0.7) + 
  geom_sf(data = pumps_4326, color = "red", size = 3)

```


# ------------------------------------------------------------------------------
