---
title: "SQL IV: Creating Databases"
format: pdf
---

# ------------------------------------------------------------------------------
# load packages and connect to database server

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
library(duckdb)
```


As an example, we will consider the Saturday Night Live datasets available on the *snldb* GitHub [repo](https://github.com/hhllcks/snldb). Data is scraped from [SNL archives](http://www.snlarchives.net) and http://www.imdb.com/title/tt0072562 by Hendrik Hilleckes and Colin Morris.

```{r}

con_snl <- DBI::dbConnect(duckdb::duckdb(), dbdir = "snl_db")

```


```{sql connection=con_snl}
SHOW TABLES;
```


# ------------------------------------------------------------------------------
# accessing data (csv files) from GitHub (or, elsewhere)

Notice that there are eleven *.csv* files available in the *output* folder. Let's look at the *casts* dataset in R, so that we understand the data we want to load into the database.

```{r}
# read in the data

casts <- readr::read_csv("")

```


```{r}
glimpse(casts)

```


```{r}
summary(casts)

```


```{r}
# save the dataset as a .csv file on your (local) computer

write_csv(casts, "casts.csv")   

```


# ------------------------------------------------------------------------------
# data types in SQL

Consult [data types](https://dev.mysql.com/doc/refman/8.0/en/data-types.html) for details.


# ------------------------------------------------------------------------------
# importing data

Once the database is set up, you will be ready to import .csv files into the database as tables. Importing .csv files as tables requires a series of steps:

* a *USE* statement that ensures we are in the right database.
* a series of *DROP TABLE* statements that drop any old tables with the same names as the ones we are going to create.
* a series of *CREATE TABLE* statements that specify the table structures and field types.
* a series of *COPY* (or *LOAD DATA* for MySQL) statements that read the data from the .csv files into the appropriate tables.


```{sql connection=con_snl}

USE snl_db;

```


```{sql connection=con_snl}

DROP TABLE IF EXISTS casts;

```


```{sql connection=con_snl}

CREATE TABLE casts (
  aid VARCHAR(255) NOT NULL DEFAULT ' ',
  sid SMALLINT NOT NULL DEFAULT 0,
  featured BOOLEAN NOT NULL DEFAULT FALSE,  /* use TINYINT(1) in MySQL instead of BOOLEAN) */
  first_epid INTEGER DEFAULT NULL,
  last_epid INTEGER DEFAULT NULL,
  update_anchor BOOLEAN NOT NULL DEFAULT FALSE,
  n_episodes SMALLINT NOT NULL DEFAULT 0,
  season_fraction DOUBLE NOT NULL DEFAULT 0,
  PRIMARY KEY (sid, aid)
);

```


```{sql connection=con_snl}

COPY casts FROM 'casts.csv' (FORMAT CSV, HEADER, NULL 'NA');

```


# ------------------------------------------------------------------------------
# checks

```{sql connection=con_snl}
SHOW DATABASES;
```


```{sql connection=con_snl}
SHOW TABLES;
```


```{sql connection=con_snl}
DESCRIBE casts;
```


```{sql connection=con_snl}
SELECT * FROM casts LIMIT 10;
```



# ------------------------------------------------------------------------------
# shortcut (not accepted for Project 2)

```{r}
# read in the data

episodes <- readr::read_csv("")

glimpse(episodes)

```


```{r}
# save the dataset as a .csv file on your (local) computer

write_csv(episodes, "episodes.csv")   

```


```{r}

duckdb_read_csv(con = con_snl, name = "episodes", files = "episodes.csv")

```


```{sql connection=con_snl}
SHOW TABLES;
```


```{sql connection=con_snl}
DESCRIBE episodes;
```


```{sql connection=con_snl}
SELECT * FROM episodes LIMIT 10;
```


# ------------------------------------------------------------------------------

```{r}
# terminate the SQL connection

DBI::dbDisconnect(con_snl, shutdown = TRUE)

```

# ------------------------------------------------------------------------------

