---
title: "Joins, Unions, and Subqueries"
execute: 
  eval: true
  warning: false
  message: false
  cache: false
---

In our previous chapters, we focused on extracting and aggregating data from a single table. However, the defining feature of a **relational database** is that data is deliberately distributed across multiple tables. This design principle, known as **normalization**, eliminates data redundancy, saves storage space, and ensures data integrity. 

For example, instead of repeating the full name of an airline on every single row of the `flights` dataset, the `flights` table only stores a brief carrier code (e.g., "DL"). If we want the full name ("Delta Air Lines Inc."), we must link the `flights` table to the `carriers` table. 

In this chapter, we explore how to stitch these normalized tables back together.

## load packages and connect to database server

Let us load our required packages and establish our connection to the `airlines` database.

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
```


```{r}
# access airlines database

con_air <- mdsr::dbConnect_scidb("airlines")

DBI::dbListTables(con_air)   # list tables in database

```


We will create our standard lazy pointers for the three tables we will analyze in this chapter: `flights`, `carriers`, and `airports`.


```{r}
# access tables from the airlines database

flights <- dplyr::tbl(con_air, "flights")

carriers <- dplyr::tbl(con_air, "carriers")

airports <- dplyr::tbl(con_air, "airports")

```


Before joining tables, it is standard practice to examine their schemas using the `DESCRIBE` command. We must identify the **keys** — the columns that exist in both tables and will serve as our connective tissue.

### `flights` table


```{sql}
--| connection: con_air

DESCRIBE flights;

```

```{sql}
--| connection: con_air

SELECT * FROM flights LIMIT 10;
```


### `carriers` table

```{sql}
--| connection: con_air

DESCRIBE carriers;

```

```{sql}
--| connection: con_air

SELECT * FROM carriers LIMIT 100;
```


### `airports` table

```{sql}
--| connection: con_air

DESCRIBE airports;

```

```{sql}
--| connection: con_air

SELECT * FROM airports LIMIT 100;
```




## Joins (Horizontal Integration)

To combine tables horizontally (adding columns from one table to another), we use `SQL` joins. In general, we must specify four pieces of information to successfully join two tables:

* name of the first table
* type of join
* name of the second table
* condition(s) of the join


The `MySQL` dialect supports several types of joins, which translate directly to the verbs we use in `dplyr`:

* **JOIN (or INNER JOIN)**: Returns only the rows that have matching values in both tables. Equivalent to `dplyr::inner_join()` in `R`.

* **LEFT JOIN** - Returns all rows from the left table, and the matched rows from the right table. Unmatched rows yield `NULL`. Equivalent to `dplyr::left_join()`.

* **RIGHT JOIN** - Returns all rows from the right table, and the matched rows from the left. Equivalent to `dplyr::right_join()`.

* **CROSS JOIN** - Returns the Cartesian product of rows from both tables. (Rarely used in standard data science workflows as it generates massive output).

::: {.callout-note title="The Missing FULL JOIN"}
Notice that `FULL JOIN` (equivalent to `R`'s `full_join()`) is missing from this list. `MySQL` and `MariaDB` do not natively support full outer joins. If we were using `PostgreSQL` or `DuckDB`, `FULL OUTER JOIN` would be available. In `MySQL`, we have to construct a full join manually by taking a `LEFT JOIN` and combining it with a `RIGHT JOIN` using a `UNION`.
:::


### `INNER JOIN`

Suppose we want to access the full airport names for all flights that flew out of Appleton (ATW) in 2013. The flights table contains the destination code (`dest`), and the airports table contains the destination code (`faa`) alongside the full name.

```{sql}
--| connection: con_air

-- access airport names for flights that flew out of Appleton in 2013

SELECT origin, dest, a.name AS dest_name, flight, carrier
FROM flights AS f
JOIN airports AS a ON f.dest = a.faa
WHERE year = 2013 AND origin = 'ATW'
LIMIT 10;


```


Notice our use of the `AS` keyword in the `FROM` and `JOIN` clauses to create aliases (`f` for `flights` and `a` for `airports`). This is a critical best practice. It saves us from typing the full table name repeatedly, and more importantly, it prevents ambiguity errors. If both tables contained a column named `year`, calling `SELECT year` would cause the query to crash; we would have to explicitly declare `SELECT f.year`.


Data architectures often require us to chain **multiple joins** together to assemble a complete picture. What if we want the full names of the origin airport, the destination airport, and the carrier?

Because we need airport names for both the origin and the destination, we must join the `airports` table to our query twice. To do this, we must give the `airports` table two distinct aliases.


```{sql}
--| connection: con_air

-- access both airport and carrier names for flights that flew out of Appleton in 2013

SELECT origin, a2.name AS origin_name, 
       dest, a1.name AS dest_name, 
       flight, 
       f.carrier, c.name AS carrier_name
FROM flights AS f
JOIN airports AS a1 ON f.dest = a1.faa
JOIN airports AS a2 ON f.origin = a2.faa
JOIN carriers AS c ON f.carrier = c.carrier
WHERE origin = 'ATW' AND year = 2013
LIMIT 10;

```


### `LEFT JOIN`

While inner joins find intersections, `LEFT JOIN` is exceptionally useful for finding missing or mismatched data.

Let us intentionally look for destination airport codes in the `flights` table that do not have a corresponding entry in the `airports` table. We do this by performing a `LEFT JOIN` and filtering for rows where the airport name resolves to `NULL`.

```{sql}
--| connection: con_air

-- using LEFT JOIN

SELECT year, month, day, origin, dest,
      a.name AS dest_name,
      flight, carrier
FROM flights AS f
LEFT JOIN airports AS a ON f.dest = a.faa
WHERE a.name IS NULL
LIMIT 10;

```


In applied contexts, queries like this are fundamental to data auditing. In this specific dataset, this query reveals flights traveling to US territories like Puerto Rico (BQN, SJU) or the Virgin Islands (STT). Because they operate slightly outside the standard domestic FAA categorization used to build our `airports` table, they show up as missing values in our join.


::: {.callout-note title="Handling Missing Values in `SQL`"}
Users of `R` are used to `NA` and functions like `is.na()`. `SQL` has a related but stricter concept: `NULL`, representing an **unknown or missing value**. The critical difference is that `NULL` cannot be compared using the standard `=` operator. A test like `a.name = NULL` does not evaluate to `TRUE` or `FALSE` — it evaluates to `UNKNOWN`, and rows where this happens are silently excluded from the result, with no error raised. To correctly test for missing values, `SQL` provides the dedicated keywords `IS NULL` and `IS NOT NULL`. 

Notice the difference in the `WHERE` clause between the queries above and below..

```{sql}
--| connection: con_air

SELECT year, month, day, origin, dest,
      a.name AS dest_name,
      flight, carrier
FROM flights AS f
LEFT JOIN airports AS a ON f.dest = a.faa
WHERE a.name = NULL
LIMIT 10;

```

:::


## Unions (Vertical Integration)

While joins add columns to our result set, set operations add rows. `UNION` allows us to append the results of one query directly beneath the results of another, much like `dplyr::bind_rows()`.

Let us extract a dataset combining all flights that flew from ATW to MSP on April 6, 2013, with all flights that flew from JFK to ORD on the same day.

```{sql}
--| connection: con_air

-- extract all flights that flew from ATW to MSP and from JFK to ORD on April 6, 2013

(SELECT year, month, day, origin, dest, flight, carrier
FROM flights
WHERE year = 2013 AND month = 4 AND day = 6 AND origin = 'ATW' AND dest = 'MSP')

UNION

(SELECT year, month, day, origin, dest, flight, carrier
FROM flights
WHERE year = 2013 AND month = 4 AND day = 6 AND origin = 'JFK' AND dest = 'ORD')

LIMIT 10;


```


::: {.callout-warning title="Union Edge Cases"}
When using `UNION`, we must be aware of two strict database rules:

* **Column Matching**: Unlike `R`'s `bind_rows()`, which politely aligns columns by their names and fills in `NA`s where they don't match, `SQL` `UNION` does not check column names. It blindly stacks the tables based on the number of columns and their data types. If query A returns a string in column 1, and query B returns an integer in column 1, the query will fail.

* **Deduplication**: By default, `UNION` removes exact duplicate rows from the final result set. Searching for and removing duplicates is computationally expensive. If we know our datasets are mutually exclusive (as they are above) or if we want to retain duplicates, we should use `UNION ALL`, which simply appends the rows and executes much faster.

:::



## Subqueries

As our analytical questions grow more sophisticated, we often encounter situations where we cannot get our answer in a single pass. We need to evaluate a query, hold the result in memory, and then query that result.

We do this using subqueries. A **subquery** is simply a complete `SELECT` statement wrapped in parentheses and placed inside the `FROM` or `WHERE` clause of a parent query.

Let us return to the `UNION` query we just wrote above. What if we don't want to see the rows, but rather we just want a count of the total number of flights across both of those specific routes? We can wrap the entire `UNION` in parentheses and run a `COUNT(*)` over it.

```{sql}
--| connection: con_air

-- count the number of flights that satisfy the criteria above

SELECT COUNT(*) AS num_flights
FROM 
((SELECT year, month, day, origin, dest, flight, carrier
FROM flights
WHERE year = 2013 AND month = 4 AND day = 6 AND origin = 'ATW' AND dest = 'MSP')
UNION
(SELECT year, month, day, origin, dest, flight, carrier
FROM flights
WHERE year = 2013 AND month = 4 AND day = 6 AND origin = 'JFK' AND dest = 'ORD')
) AS temp_subquery;

```

::: {.callout-tip title="The Mandatory Alias"}
When we place a subquery inside a `FROM` clause, we are essentially creating a temporary table on the fly (often called a **derived table**). In `MySQL` and `MariaDB`, every derived table must have its own alias, even if we never explicitly refer to that alias in the `SELECT` statement. Notice the `AS temp_subquery` at the very end of our code block; without it, the `SQL` engine would throw a syntax error.
:::


## terminate connection

As always, we conclude our workflow by responsibly closing our database connections to free up memory on both our local machine and the remote server.

```{r}
# terminate the SQL connection

DBI::dbDisconnect(con_air, shutdown = TRUE)

```



