---
title: "Accessing Databases and Database Tables"
execute: 
  eval: true
  warning: false
  message: false
  cache: false
---



## load packages

To interact with a database using `R`, we need a specific set of tools. The `tidyverse` collection of `R` packages provides our standard data manipulation framework, while `DBI` (Database Interface) serves as the common architecture for communicating with **Relational DataBase Management Systems (RDBMS)**. Because different databases (`PostgreSQL`, `MySQL`, `SQLite`) speak slightly different dialects of `SQL`, we also need a specific driver. In this case, we use `RMariaDB` to communicate with `MySQL/MariaDB` servers. Finally, `dbplyr` is the critical translation layer that allows us to write `dplyr` code that `R` automatically converts into `SQL`.

```{r}
library(tidyverse)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
```



## databases and accessing them

Relational databases operate on a **client-server** model. In this architecture, the data does not live on our local machine. Instead, it is housed on a centralized remote server. Our `R` session (the client) must send a formal request over a network to access that data.

We initialize this connection using the `mdsr` package, which provides a convenient wrapper function to connect to a public, read-only educational database.

Here, `con_air` is the connection to the `airlines` database on the `scidb` server located on Amazon Web Services. Check out more details on the help page (`?mdsr::dbConnect_scidb`).

```{r}
# connecting to the airlines database

con_air <- mdsr::dbConnect_scidb("airlines")

# remember the name of the connection

```

::: {.callout-note title="The Client-Server Analogy"}
Think of the **client-server** model like a restaurant. 
We (the client) sit at a table and look at the menu. We do not go into the kitchen to cook the food ourself. Instead, we send our order (the query) via a waiter (the connection) to the kitchen (the server). The kitchen processes the heavy lifting and sends back only our specific meal (the results).
:::


While wrapper functions like the one above are convenient for teaching, they hide the underlying mechanics. In an industry setting, connecting to a database requires explicit routing and authentication parameters. We must specify the exact server location, the name of the database, our username, and our password.

Here is what an explicit connection via `DBI` looks like:

```{r}
#| eval: false
# access wai database

con_wai <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "wai",
  host = "scidb.smith.edu",
  user = "waiuser",
  password = "smith_waiDB")

```

::: {.callout-warning title="Password Security"}
It is not ideal and hence, not recommended, to share passwords. The code chunk above is shown only for teaching purposes. 
:::

## database tables and accessing them

Once a connection is established, the first step is typically reconnaissance: we need to know what tables exist within the database.

We can achieve this using `R` syntax via the `DBI` package:

```{r}
# lists tables in the airlines database

DBI::dbListTables(con_air)   

```

Alternatively, because Quarto is a polyglot authoring framework, we can write raw `SQL` directly. By specifying `connection = "con_air"` or using `#| connection: con_air` in the chunk options, Quarto knows to pass the code inside the chunk directly to the database rather than to the `R` interpreter.

Note, the following is a `SQL` code chunk.

```{sql}
--| connection: con_air

-- lists tables in the database

SHOW TABLES;

```

::: {.callout-note title="SQL Syntax Conventions"}
Notice that we wrote SHOW TABLES in all capital letters and ended the statement with a semicolon (;). While `SQL` is technically case-insensitive, capitalizing `SQL` keywords is a universal standard that drastically improves readability, distinguishing commands from table and column names. The semicolon acts as a terminator, signaling to the server that the query is complete and ready for execution.
:::


Knowing the tables exist is not enough; we must establish a reference to a specific table to query it. We do this using `dplyr::tbl()`. This creates a lazy reference to a remote database table without downloading it. Operations on this object are translated to `SQL` and executed on the server only when results are actually needed.

Here, we query the `flights` table from the established connection to the `airlines` database.


```{r}
# reference to flights table from the airlines database

flights <- dplyr::tbl(con_air, "flights")

```

::: {.callout-note title="Object Distinctions"}
It's easy to conflate the objects we have created thus far. Notice the distinction between `con_air` (the connection object), `airlines` (the database), and `flights` (the table reference). The following hierarchy would be helpful

$$\text{server} \rightarrow \text{database} \rightarrow \text{table} \rightarrow \text{table reference}$$
:::


Notice the output of `class(flights)`. It is not a standard `data.frame` or `tbl_df`. It is a `tbl_dbi` and `tbl_sql`. This distinction is profound because it introduces the concept of lazy evaluation. 

```{r}
# inspect the object type

class(flights)

```


When we print the `flights` object below, it provides a quick glimpse of the columns and the first few rows. However, we will notice it says `??` for the total number of rows. This is because the data is not actually in R's memory. `R` has merely established a pointer to the table on the server. Because counting tens of millions of rows takes computational time, the database engine skips that step until explicitly commanded otherwise. It is **"lazy"**—it does as little work as possible.


```{r}
# attempting to print the flights object

flights

```


::: {.callout-note title="Lazy `tbl`"}
Notice that typical exploratory functions such as `str` and `glimpse` would work differently with the `flights` object compared to datasets loaded in the local session memory.

```{r}
str(flights)
```
```{r}
glimpse(flights)  # notice the ?? on the number of rows
```
:::


To understand the architecture of the table without pulling data, we can use the `DESCRIBE` command in `SQL`. This returns column/field names, data types, and constraints for a table. This is a recommended first step before working with the table.




```{sql}
--| connection: con_air

DESCRIBE flights;

```

This output is conceptually similar to R's `glimpse()` function, but it exposes the underlying database types (e.g., `varchar`, `int`, `datetime`). Databases require strict typing to optimize physical storage on disk. A column defined as `varchar(3)` can only ever hold text strings up to three characters long, preventing data corruption and saving immense amounts of storage space.

Remember, earlier we were trying to figure out if it would be more efficient to store student names versus their IDs!


To fully appreciate the power of lazy evaluation, we can compare the memory footprint of a database pointer versus a materialized in-memory object in `R`. Let us create a reference to the `carriers` table.

```{r}
# reference to carriers table from the airlines database

carriers <- dplyr::tbl(con_air, "carriers")

```


```{sql}
--| connection: con_air

DESCRIBE carriers;

```


Let's check how much RAM this connection occupies in our local environment:

```{r}
# carriers lives in the SQL database and is linked remotely

carriers |>
  utils::object.size() |>   
  print(units = "Kb")


```

The object size is remarkably small (often just a few kilobytes). It only stores the connection string and metadata, not the data itself.

However, in a typical data science workflow, we eventually need the data in our local environment—usually after we have aggregated or filtered it—so we can pass it to local functions like `ggplot2` for visualization. To bridge the gap between the remote database and our local `R` session, we use the `collect()` function.

```{r}

carriers_df <- carriers |>  dplyr::collect()   

```


Let's check the memory size now:

```{r}
# carriers lives in memory in R as carriers_df

carriers_df |>
  utils::object.size() |>
  print(units = "Kb")

```

The materialized data frame is significantly larger because the actual data now resides in our computer's RAM.

::: {.callout-warning title="The Danger of `collect()`"}
We should never run `collect()` on a massive, unaggregated table like `flights`. That table contains hundreds of millions of rows. If we attempt to `collect()` the entire table, we will exhaust our machine's RAM, and our `R` session will crash. The proper workflow is to use `dbplyr` or `SQL` to filter and summarize the data on the remote server, and only `collect()` the finalized, summarized results into R.

Another safer alternative (if needed) is

```{r}
# flights_df lives in memory in R

flights_df <- flights |> 
  head(5) |> 
  collect()

```

:::

## terminate connection

Just as we would close a file after editing it, we must explicitly sever our connection to a remote database when our analysis is complete. Open connections consume resources on both our local machine and the remote server.

```{r}
# terminate the SQL connection

DBI::dbDisconnect(con_air, shutdown = TRUE)

```


Terminating connections is an essential habit for responsible computing, particularly when sharing server resources in an industry or academic environment.





