---
title: "Writing SQL queries"
execute: 
  eval: true
  warning: false
  message: false
  cache: false
---

In our previous chapter, we established how to connect `R` to a remote relational database. However, simply connecting to a database is only the first step. Our primary goal is to extract, filter, and aggregate data to answer analytical questions. 

Because we are operating within the `R` ecosystem but interacting with a `SQL` database, we have three distinct methods for sending instructions to the server:

1.  Translating `R`'s `dplyr` syntax into `SQL` using the `dbplyr` package.
2.  Passing raw `SQL` strings through `R` functions using the `DBI` package.
3.  Writing native `SQL` directly within Quarto code chunks.

In this chapter, we will explore all three methodologies, gradually transitioning from the familiar `tidyverse` syntax to writing robust, native `SQL`.


## load packages and connect to database server

We begin by loading our required packages and establishing our connection to the `airlines` database, just as we did previously. We will also create a reference pointer to the `flights` table, which contains millions of commercial flight records.


```{r}
library(tidyverse)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
```


```{r}
# connect to the airlines database

con_air <- mdsr::dbConnect_scidb("airlines")

DBI::dbListTables(con_air)   # list tables in database

```


```{r}
# reference to flights table from the airlines database

flights <- dplyr::tbl(con_air, "flights")

```


```{sql}
--| connection: con_air

DESCRIBE flights;

```




## SQL in R: Part 1 - the 'dbplyr' R package will directly translate 'dplyr' code into SQL

For data scientists already comfortable with the `tidyverse`, the `dbplyr` package is an extraordinary tool. It acts as a translation engine, taking our standard `dplyr` verbs (`filter`, `arrange`, `select`, `mutate`, `summarize`, `group_by`) and silently converting them into the specific `SQL` dialect required by our database (in this case, `MariaDB/MySQL`).

Suppose we want to find the minimum and maximum year recorded in our `flights` table. We can write standard `dplyr` code:

```{r}

yrs <- flights |>
  summarize(min_year = min(year, na.rm = TRUE), 
            max_year = max(year, na.rm = TRUE))

yrs

```

The result `yrs` is a **lazy** query — it holds the `SQL` definition, not the data itself.

The data is processed on the remote server, and only the summarized one-row result is returned to our console. But how did `R` communicate this to the database? We can peek under the hood using the `show_query()` function. This reveals the SQL that `dbplyr` generated under the hood.

```{r}
dplyr::show_query(yrs)  
```

::: {.callout-tip title="The SQL Vocabulary"}
Looking at the output of `show_query()`, we see several foundational `SQL` keywords:

* `SELECT`: Determines which columns or calculations we want to retrieve.

* `AS`: Creates an alias, similar to the = sign in a `mutate()` or `summarize()` function.

* `FROM`: Specifies the table we are querying.

Notice also that the query ends with a semicolon (;).
:::

Let us try a more complex example. We want to extract all flights from 2013 that traveled between Los Angeles (LAX) and Boston (BOS).

```{r}

la_bos <- flights |>
  filter(year == 2013 & 
           ((origin == "LAX" & dest == "BOS") | 
           (origin == "BOS" & dest == "LAX"))) 

la_bos

```

When we inspect the `SQL` translation for this filter, we notice some critical syntactical differences between `R` and `SQL`.

```{r}
dplyr::show_query(la_bos)
```

In the `SQL` translation, our `R` `filter()` statement becomes a `SQL` `WHERE` clause. Furthermore, notice that `SQL` uses a single = for logical evaluation, whereas `R` requires ==. Finally, our symbolic operators `&` and `|` are translated into explicit `AND` and `OR` keywords.

Once we are satisfied with our query and wish to bring the data into our local R environment for visualization or modeling, we append the `collect()` function. This should be done only when the result set is small enough to fit in memory.


```{r}

la_bos <- la_bos |>
  collect()

la_bos

```


::: {.callout-warning title="Where `dbplyr` translation breaks down"}

Notice the code below. Depending on the database backend, this either fails outright or produces a warning, because `cumsum()` requires a `SQL` **window function**, and not every `dplyr` function has a clean, universal `SQL` equivalent — especially once we leave basic aggregation and move into ordered, row-by-row calculations. This is precisely the kind of situation where dropping down to raw `SQL` becomes necessary rather than a stylistic choice. In the sections that follow, we introduce two ways to do exactly that.

```{r}
#| warning: true

# attempting a window function (running total) using dplyr syntax

dbplyr_fail <- flights |>
  filter(origin == "ATW") |>
  group_by(year) |>
  summarize(num_flights = n()) |>
  mutate(running_total = cumsum(num_flights))


dplyr::show_query(dbplyr_fail)

```

:::

## SQL in R: Part 2 - using the DBI R package to write SQL queries in r code chunks

While `dbplyr` is incredibly convenient, it cannot translate every possible `SQL` function, especially highly specialized window functions or dialect-specific commands. In these instances, we can bypass the translation layer and pass raw `SQL` strings directly to the database using `DBI::dbGetQuery()`.

```{r}
# retrieves the first 8 rows from flights

DBI::dbGetQuery(con_air,
                "SELECT * FROM flights LIMIT 8;")  

```

In `SQL`, the `*` symbol is a wildcard meaning "all columns." The `LIMIT` clause is the `SQL` equivalent of `R`'s `head()` function, restricting the output size.

We can also perform complex aggregations directly within the string. Here, we count the number of flights per year and order the results.

```{r}

DBI::dbGetQuery(con_air, 
  "SELECT year, count(*) AS num_flights FROM flights GROUP BY year ORDER BY num_flights;")

```



## SQL in R: Part 3 - writing SQL queries in sql code chunks

Passing `SQL` strings inside `R` functions can become visually cumbersome for long, multi-line queries. Because Quarto is a polyglot environment, we can change the code chunk type from `{r}` to `{sql}`. This provides proper syntax highlighting and allows us to write native `SQL` exactly as we would in a dedicated database client.

Let us recreate our previous `dbplyr` queries using native `SQL` chunks.

```{sql}
--| connection: con_air

-- display minimum and maximum year

SELECT MIN(year) AS min_year, 
       MAX(year) AS max_year
FROM flights;

```


```{sql}
--| connection: con_air

-- extract flights between LAX and BOS in 2013

SELECT *
FROM flights
WHERE (year = 2013 AND 
      ((origin = 'LAX' AND dest = 'BOS') OR 
      (origin = 'BOS' AND dest = 'LAX')));
      
-- notice the use of = (not ==) for equality comparisons

```


`LIMIT` restricts the number of rows returned. This is useful for previewing data without pulling the entire table from the server.


```{sql}
--| connection: con_air

-- select all columns and limit to top 2 rows

SELECT * 
FROM flights 
LIMIT 2;

```


`GROUP BY` partitions rows into groups sharing the same value of one or more columns, and aggregate functions like `COUNT()` are applied within each group. `ORDER BY` sorts the result; the default sort order is ascending (use `ORDER BY num_flights DESC` for descending order).


```{sql}
--| connection: con_air

-- select year, and the computing num_flights from the flights table 
-- grouping by year
-- ordering by the num_flights

SELECT year, 
       COUNT(*) AS num_flights 
FROM flights 
GROUP BY year 
ORDER BY num_flights;

```


::: {.callout-note title="Written order of `SQL` clauses"}

**Queries** in `SQL` start with the **SELECT** keyword and consist of several **clauses**, which have to be written in this order:

* **SELECT** allows you to list the columns, or functions operating on columns, that you want to retrieve. 

* **FROM** specifies the table where the data are.

* **JOIN** allows you to stitch together two or more tables using a key. 

* **WHERE** allows you to extract rows according to some criteria. 

* **GROUP BY** allows you to aggregate the records according to some shared value. 

* **HAVING** is like a `WHERE` clause that operates on the result set—not the original rows themselves. 

* **ORDER BY** orders the rows of the result set. 

* **LIMIT** restricts the number of rows in the output. 

:::


The following queries explore the `flights` table, starting with basic retrieval and building toward more complex aggregations.

```{sql}
--| connection: con_air

DESCRIBE flights;

```


### query 1

`SELECT *` retrieves all columns. Combined with `LIMIT`, this is a quick way to preview the contents of a table without loading all rows.

```{sql}
--| connection: con_air

-- display the first ten rows of the flights table

SELECT *
FROM flights
LIMIT 10;

```

Essentially, `LIMIT` takes two arguments - the first skips a specific number of rows (offset), the second returns the specified row count after skipping.

```{sql}
--| connection: con_air

-- skip the first 18,008,362 rows and return the next 10
-- how did we find out the total number of rows? 

SELECT *
FROM flights
LIMIT 18008362, 10;

```


### query 2

Rather than retrieving all columns, we can name specific columns in the `SELECT` clause. This reduces the amount of data transferred from the server and makes results easier to interpret.

```{sql}
--| connection: con_air

-- display year, month, day, origin, dep_time, sched_dep_time, dep_delay for the first ten rows of the flights table

SELECT year, month, day, origin, dep_time, sched_dep_time, dep_delay
FROM flights
LIMIT 10;

```


We can also concatenate strings and cast them into date formats natively.

```{sql}
--| connection: con_air

SELECT STR_TO_DATE(CONCAT(year, '-', month, '-', day), '%Y-%m-%d') AS 'date', origin, dep_time, sched_dep_time, dep_delay
FROM flights
LIMIT 10;



```

### query 3

`COUNT(*)` counts every row in the table regardless of `NULL` values in individual columns. Aliasing the result with `AS` gives the output column a descriptive name.

```{sql}
--| connection: con_air

-- count the total number of rows in the flights table

SELECT COUNT(*) AS total_rows
FROM flights;

```


`COUNT(*)` counts *rows*, but we often want to count **unique values** instead — for example, how many different carriers are in this table. In `R`, `n_distinct()` does this; in `SQL`, we pair `COUNT()` with the `DISTINCT` keyword. 

```{sql}
--| connection: con_air

SELECT COUNT(DISTINCT carrier) AS num_carriers
FROM flights;

```


### query 4

`WHERE` filters rows before any aggregation takes place. Here we restrict to flights from ATW to ORD in 2013, combining multiple conditions with `AND`.

```{sql}
--| connection: con_air

-- extract all info about flights that flew from Appleton (ATW) to Chicago (ORD) in 2013

SELECT *
FROM flights
WHERE origin = 'ATW' AND dest = 'ORD' AND year = 2013;

```

::: {.callout-note title="Logical operators `AND` and `OR`"}

Logical operators in `SQL` evaluate in a strict order: `AND` is evaluated before `OR`.

Imagine we are looking at our `flights` table and we want to find all flights in the year 2013 that either departed from Appleton (ATW) or arrived in Chicago (ORD).

It is tempting to write the query exactly as we speak it:

```{sql}
--| connection: con_air

SELECT COUNT(*) AS num_flights
FROM flights
WHERE year = 2013 AND origin = 'ATW' OR dest = 'ORD';

```

Because `AND` has higher precedence, the `SQL` engine secretly groups the query like this:

```{sql}
--| connection: con_air

SELECT COUNT(*)
FROM flights
WHERE (year = 2013 AND origin = 'ATW') OR dest = 'ORD';

```

What this actually returns:

1. Every flight that departed from ATW in 2013.

2. PLUS every single flight that landed in ORD, regardless of the year, and regardless of where it came from.

Instead of returning a small subset of specific 2013 flights, our query returns millions of rows, pulling in ORD flights from 2001, 2010, etc.

To fix this, we use parentheses to force the database to evaluate the `OR` condition as a single, combined logical unit before it evaluates the `AND` condition.

```{sql}
--| connection: con_air

SELECT COUNT(*) AS num_flights
FROM flights
WHERE year = 2013 AND (origin = 'ATW' OR dest = 'ORD');

```

How `SQL` evaluates this:

1. First, it looks inside the parentheses: "Is the origin ATW, `OR` is the destination ORD?" (Returns True or False for each row).

2. Then, it evaluates the `AND`: "Is the year 2013 `AND` did the previous step return True?"

By wrapping the `OR` statements in brackets, we ensure that the `year = 2013` condition acts as a strict filter applied to both the origin and the destination checks.

A Golden Rule for `SQL`:

**Whenever you have a `WHERE` clause that mixes `AND` and `OR`, always use parentheses. Even if you memorize the precedence rules, parentheses make your code instantly readable to your colleagues and future self, eliminating any ambiguity about your analytical intentions.**

:::


::: {.callout-note title="`BETWEEN` vs `IN`"}

Notice the difference between the `BETWEEN` and `IN` operators.


```{sql}
--| connection: con_air

-- number of flights that were delayed at departure by less than 10 minutes

SELECT COUNT(*) as num_rows
FROM flights
WHERE dep_delay BETWEEN 0 AND 10;

```

```{sql}
--| connection: con_air

-- number of flights that were delayed by either 0 or 10 minutes

SELECT COUNT(*) as num_rows
FROM flights
WHERE dep_delay IN (0,10);



```

:::


<!-- ::: {.callout-note title="Handling Missing Values in `SQL`"} -->
<!-- Users of `R` are used to `NA` and functions like `is.na()`. `SQL` has a related but stricter concept: `NULL`, representing an unknown or missing value. The critical difference is that `NULL` cannot be compared using the standard `=` operator. A test like `dep_delay = NULL` does not evaluate to `TRUE` or `FALSE` — it evaluates to `UNKNOWN`, and rows where this happens are silently excluded from the result, with no error raised. To correctly test for missing values, `SQL` provides the dedicated keywords `IS NULL` and `IS NOT NULL`. Notice both the queries below and compare `num_missing` to `num_missing_wrong` — the second will return zero, which is a common and quiet source of bugs. -->

<!-- ```{sql} -->
<!-- --| connection: con_air -->

<!-- -- flights with a missing (NULL) departure delay -->
<!-- -- these are often cancelled or diverted flights with no recorded time -->

<!-- SELECT * -->
<!-- FROM flights -->
<!-- WHERE dep_delay IS NULL; -->

<!-- ``` -->

<!-- ```{sql} -->
<!-- --| connection: con_air -->

<!-- -- this query looks like it should find the same rows, but it does not -->

<!-- SELECT COUNT(*) AS num_missing_wrong -->
<!-- FROM flights -->
<!-- WHERE dep_delay = NULL; -->

<!-- ``` -->
<!-- ::: -->


<!-- ::: {.callout-warning title="Operator Precedence Edge Case"} -->
<!-- Logical operators in `SQL` evaluate in a strict order: `AND` is evaluated before `OR`. Consider the following query meant to find flights out of Appleton (ATW) going to Chicago (ORD) in 2013: -->

<!-- ```{sql} -->
<!-- --| connection: con_air -->
<!-- --| eval: false -->

<!-- -- extract all info about flights that flew from Appleton (ATW) to Chicago (ORD) in 2013 -->

<!-- SELECT * -->
<!-- FROM flights -->
<!-- WHERE year = 2013 AND origin = 'ATW' OR dest = 'ORD'; -->


<!-- ``` -->

<!-- Because `AND` binds tighter, `SQL` interprets this as: "Find flights from ATW in 2013, `OR` find any flight going to ORD from any year and any origin." This will return millions of incorrect rows. We must always use parentheses to explicitly define our logic, just as we did in our LAX-BOS query earlier. -->
<!-- ::: -->


### query 5

Adding `GROUP BY year` counts the number of flights that operated on the ATW–ORD route per year.

`ORDER BY year` sorts the results chronologically.

```{sql}
--| connection: con_air

-- count the number of flights from Appleton (ATW) for each year

SELECT year, COUNT(*) AS num_flights
FROM flights
WHERE origin = 'ATW' AND dest = 'ORD'
GROUP BY year
ORDER BY year;

```



We can group by multiple variables by separating them with commas. Grouping by both `carrier` and `year` lets us see how many flights each airline operated on the ATW–ORD route per year.

```{sql}
--| connection: con_air

-- which carriers flew from Appleton (ATW) to Chicago (ORD)
-- how many flights did they do per year

SELECT carrier, year, COUNT(*) AS num_flights
FROM flights 
WHERE origin = 'ATW' AND dest = 'ORD'
GROUP BY carrier, year
ORDER BY carrier, year;



```

### query 6

`MAX(dep_time)` within each carrier group returns the latest recorded departure time for that carrier on the ATW–ORD route.

Note that departure times in this table are stored as integers in `HHMM` format (e.g., 1435 = 2:35 PM).

```{sql}
--| connection: con_air

-- for flights that flew from Appleton (ATW) to Chicago (ORD)
-- what was the latest flight for each carrier

SELECT carrier, MAX(dep_time) AS latest_dep_time 
FROM flights 
WHERE origin = 'ATW' AND dest = 'ORD'
GROUP BY carrier
LIMIT 1;



```


### query 7

Understanding the difference between `WHERE` and `HAVING` is one of the most important hurdles in learning `SQL`. Both act as filters, but they operate at different stages of the execution pipeline.

`WHERE` filters records in the original dataset before any grouping occurs. `HAVING` filters the results of our aggregations after the `GROUP BY` clause has executed.

First, let us find the destination with the lowest average arrival delay from Appleton, using `WHERE` to filter the origin.

```{sql}
--| connection: con_air

-- for flights that flew from Appleton (ATW) 
-- which destination had the lowest average arrival delay time

SELECT dest, COUNT(*) AS num_flights, AVG(arr_delay) AS avg_arr_delay
FROM flights 
WHERE origin = 'ATW'
GROUP BY dest
ORDER BY avg_arr_delay;


```


Suppose we want to restrict our analysis to robust routes — say, those with at least 2,000 flights. We cannot put `num_flights >= 2000` in the `WHERE` clause because `num_flights` does not exist in the original table; it is an aggregated calculation. Therefore, we must use `HAVING`.

```{sql}
--| connection: con_air

-- for flights that flew from Appleton (ATW) 
-- consider only destinations with at least 2000 flights from Appleton (ATW)
-- which destination had the lowest average arrival delay time

SELECT dest, COUNT(*) AS num_flights, AVG(arr_delay) AS avg_arr_delay
FROM flights 
WHERE origin = 'ATW'
GROUP BY dest
HAVING num_flights >= 2000
ORDER BY avg_arr_delay;



```


`HAVING` filters groups **after** aggregation, unlike `WHERE` which filters rows **before** aggregation.


### query 8

We now have all the pieces to write a complete query using every clause introduced in this chapter. Notice how each clause does a distinct job, and how the order we write them in matches the order listed at the start of this section: we filter raw rows first (`WHERE`), then aggregate (`GROUP BY`), then filter the aggregated results (`HAVING`), then sort (`ORDER BY`), and finally restrict the output size (`LIMIT`). Try reading the query out loud, clause by clause — it should read almost like a sentence describing the question we are asking of the data.


```{sql}
--| connection: con_air

-- for flights out of Appleton (ATW) in 2013 or later
-- consider only destinations with at least 500 flights
-- find the 5 destinations with the highest average departure delay

SELECT dest,
       COUNT(*) AS num_flights,
       AVG(dep_delay) AS avg_dep_delay
FROM flights
WHERE origin = 'ATW' AND year >= 2013
GROUP BY dest
HAVING num_flights >= 500
ORDER BY avg_dep_delay DESC
LIMIT 5;

```


::: {.callout-note title="Which approach to use?"}

Having now seen all three approaches to writing `SQL` from within `R`, a natural question is: which one should we actually use? There is no single right answer, but the following rough guide is a useful starting point:

- **Use `dbplyr` (Part 1)** when your query is a fairly standard `dplyr` pipeline — filtering, grouping, summarizing — and you want your code to stay readable and consistent with the rest of a `tidyverse`-based analysis. This is the best default for most exploratory work.

- **Use `DBI::dbGetQuery()` (Part 2)** when you need a `SQL` feature `dbplyr` cannot translate (such as window functions), but the query is short enough to comfortably live as a string inside an `R` script or function.

- **Use native `{sql}` code chunks (Part 3)** when writing long, complex, or highly `SQL`-specific queries — window functions, subqueries, or multi-table joins — where syntax highlighting and formatting readability matter, or when you want a query that could be copy-pasted directly into a standalone database client.

In practice, many real analyses mix all three approaches within the same project, choosing whichever tool best fits the query at hand.

:::


## calling `SQL` generated datasets into R for data viz

Our ultimate workflow often involves doing the heavy lifting (filtering and aggregating) in `SQL`, and then pulling the summarized dataset into `R` to leverage packages like `ggplot2`. We have three ways to save the output of a `SQL` query into an `R` object:

### Method 1: using `DBI::dbGetQuery()`

```{r}
df1 <- DBI::dbGetQuery(con_air, "SELECT * FROM flights LIMIT 10")

# always check the number of rows

```

### Method 2: using `dbplyr` with raw `SQL`

```{r}

df2 <- dplyr::tbl(con_air, dplyr::sql("SELECT * FROM flights LIMIT 10")) |> dplyr::collect()

# doesn't need a semicolon

```

### Method 3: Quarto chunk options

We can write a native `SQL` chunk and use the `--| output.var: "df"` code chunk option to assign the result to an object in our `R` environment.

```{sql}
--| connection: con_air
--| output.var: "df3"


SELECT * FROM flights LIMIT 10;

```

All three methods are valid. We select the one that best fits our workflow and the complexity of the query we need to execute.

## terminate connection

Finally, always remember to disconnect your connection to the server.

```{r}
# terminate the SQL connection

DBI::dbDisconnect(con_air, shutdown = TRUE)

```





