---
title: "SQL II: Writing SQL queries"
format: pdf
---

# ------------------------------------------------------------------------------
# load packages and connect to database server

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
```


```{r}
# access airlines database

con_air <- mdsr::dbConnect_scidb("airlines")

DBI::dbListTables(con_air)   # list tables in database

```


```{r}
# access flights table from the airlines database

flights <- dplyr::tbl(con_air, "flights")

```


```{sql}
#| connection: con_air

DESCRIBE flights;

```



# ------------------------------------------------------------------------------
# SQL in R: Part 1 - the 'dbplyr' R package will directly translate 'dplyr' code into SQL

```{r}

yrs <- flights |>
  summarize(min_year = min(year, na.rm = TRUE), 
            max_year = max(year, na.rm = TRUE))

yrs

```

```{r}
dplyr::show_query(yrs)  
```



```{r}

la_bos <- flights |>
  filter(year == 2013 & 
           ((origin == "LAX" & dest == "BOS") | 
           (origin == "BOS" & dest == "LAX"))) 

la_bos

```

```{r}
dplyr::show_query(la_bos)
```

```{r}

la_bos <- la_bos |>
  collect()

la_bos

```



```{r}
# terminate the SQL connection

DBI::dbDisconnect(con_air, shutdown = TRUE)

```



# ------------------------------------------------------------------------------
# SQL in R: Part 2 - using the DBI R package to write SQL queries in r code chunks

```{r}

DBI::dbGetQuery(con_air,
                "SELECT * FROM flights LIMIT 8;")  

```


```{r}

DBI::dbGetQuery(con_air, 
  "SELECT year, count(*) AS num_flights FROM flights GROUP BY year ORDER BY num_flights;")

```


```{r}
# terminate the SQL connection

DBI::dbDisconnect(con_air, shutdown = TRUE)

```



# ------------------------------------------------------------------------------
# SQL in R: Part 3 - writing SQL queries in sql code chunks


```{sql}
#| connection: con_air

SELECT MIN(year) AS min_year, 
       MAX(year) AS max_year
FROM flights;

```


```{sql}
#| connection: con_air

SELECT *
FROM flights
WHERE (year = 2013 AND 
      ((origin = 'LAX' AND dest = 'BOS') OR 
      (origin = 'BOS' AND dest = 'LAX')));

```


```{sql}
#| connection: con_air

SELECT * 
FROM flights 
LIMIT 2;

```


```{sql}
#| connection: con_air

SELECT year, 
       COUNT(*) AS num_flights 
FROM flights 
GROUP BY year 
ORDER BY num_flights;

```


Queries in SQL start with the *SELECT* keyword and consist of several clauses, which have to be written in this order:

* *SELECT* allows you to list the columns, or functions operating on columns, that you want to retrieve. 

* *FROM* specifies the table where the data are.

* *JOIN* allows you to stitch together two or more tables using a key. 

* *WHERE* allows you to extract rows according to some criteria. 

* *GROUP BY* allows you to aggregate the records according to some shared value. 

* *HAVING* is like a WHERE clause that operates on the result set—not the original rows themselves. 

* *ORDER BY* orders the rows of the result set. 

* *LIMIT* restricts the number of rows in the output. 


```{sql}
#| connection: con_air

DESCRIBE flights;

```



```{sql}
#| connection: con_air

# display the first ten rows of the flights table



```


```{sql}
#| connection: con_air

# display year, month, day, origin, dep_time, sched_dep_time, dep_delay for the first ten rows of the flights table



```


```{sql}
#| connection: con_air

# count the total number of rows in the flights table



```


```{sql}
#| connection: con_air

# extract all info about flights that flew from Appleton (ATW) to Chicago (ORD) in 2013



```


```{sql}
#| connection: con_air

# count the number of flights from Appleton (ATW) for each year



```


```{sql}
#| connection: con_air

# which carriers flew from Appleton (ATW) to Chicago (ORD)
# how many flights did they do per year



```


```{sql}
#| connection: con_air

# for flights that flew from Appleton (ATW) to Chicago (ORD)
# what was the latest flight for each carrier



```


```{sql}
#| connection: con_air

# for flights that flew from Appleton (ATW) 
# which destination had the lowest average arrival delay time



```


```{sql}
#| connection: con_air

# for flights that flew from Appleton (ATW) 
# consider only destinations with at least 2000 flights from Appleton (ATW)
# which destination had the lowest average arrival delay time



```


```{r}
# terminate the SQL connection

DBI::dbDisconnect(con_air, shutdown = TRUE)

```


# ------------------------------------------------------------------------------


