---
title: "Joins, Unions, and Subqueries"
execute: 
  eval: true
  warning: false
  message: false
  cache: false
---

In our previous chapters, we focused on extracting and aggregating data from a single table. However, the defining feature of a **relational database** is that data is deliberately distributed across multiple tables. This design principle, known as **normalization**, eliminates data redundancy, saves storage space, and ensures data integrity. 

For example, instead of repeating the full name of an airline on every single row of the `flights` dataset, the `flights` table only stores a brief carrier code (e.g., "DL"). If we want the full name ("Delta Air Lines Inc."), we must link the `flights` table to the `carriers` table. 

In this chapter, we explore how to stitch these normalized tables back together.

## load packages and connect to database server

Let us load our required packages and establish our connection to the `airlines` database.

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
```


```{r}
# access airlines database

con_air <- mdsr::dbConnect_scidb("airlines")

DBI::dbListTables(con_air)   # list tables in database

```


We will create our standard lazy pointers for the three tables we will analyze in this chapter: `flights`, `carriers`, and `airports`.


```{r}
# access tables from the airlines database

flights <- dplyr::tbl(con_air, "flights")

carriers <- dplyr::tbl(con_air, "carriers")

airports <- dplyr::tbl(con_air, "airports")

```


Before joining tables, it is standard practice to examine their schemas using the `DESCRIBE` command. We must identify the **keys** — the columns that exist in both tables and will serve as our connective tissue.

### `flights` table


```{sql}
--| connection: con_air

DESCRIBE flights;

```

```{sql}
--| connection: con_air

SELECT * FROM flights LIMIT 10;
```


### `carriers` table

```{sql}
--| connection: con_air

DESCRIBE carriers;

```

```{sql}
--| connection: con_air

SELECT * FROM carriers LIMIT 100;
```


### `airports` table

```{sql}
--| connection: con_air

DESCRIBE airports;

```

```{sql}
--| connection: con_air

SELECT * FROM airports LIMIT 100;
```




## Joins (Horizontal Integration)

To combine tables horizontally (adding columns from one table to another), we use `SQL` joins. In general, we must specify four pieces of information to successfully join two tables:

* name of the first table
* type of join
* name of the second table
* condition(s) of the join


The `MySQL` dialect supports several types of joins, which translate directly to the verbs we use in `dplyr`:

* **JOIN (or INNER JOIN)**: Returns only the rows that have matching values in both tables. Equivalent to `dplyr::inner_join()` in `R`.

* **LEFT JOIN** - Returns all rows from the left table, and the matched rows from the right table. Unmatched rows yield `NULL`. Equivalent to `dplyr::left_join()`.

* **RIGHT JOIN** - Returns all rows from the right table, and the matched rows from the left. Equivalent to `dplyr::right_join()`.

* **CROSS JOIN** - Returns the Cartesian product of rows from both tables. (Rarely used in standard data science workflows as it generates massive output).

::: {.callout-note title="The Missing FULL JOIN"}
Notice that `FULL JOIN` (equivalent to `R`'s `full_join()`) is missing from this list. `MySQL` and `MariaDB` do not natively support full outer joins. If we were using `PostgreSQL` or `DuckDB`, `FULL OUTER JOIN` would be available. In `MySQL`, we have to construct a full join manually by taking a `LEFT JOIN` and combining it with a `RIGHT JOIN` using a `UNION`.
:::


### `INNER JOIN`

Suppose we want to access the full airport names for all flights that flew out of Appleton (ATW) in 2013. The flights table contains the destination code (`dest`), and the airports table contains the destination code (`faa`) alongside the full name.

Before we write our first join, it is worth noting a small but important habit: once a query involves more than one table, we typically give each table a short **alias** (e.g., `f` for `flights`, `a` for `airports` below) using the `AS` keyword. This is not just a convenience for shorter typing — it prevents **ambiguity errors**. If two tables in a join both contain a column with the same name (such as `year`), simply writing `SELECT year` leaves the database unable to determine which table's `year` we mean, and the query will fail. Prefixing the column with its table alias, as in `f.year`, resolves the ambiguity. We will see this pattern in every join query below.


```{sql}
--| connection: con_air

-- access airport names for flights that flew out of Appleton in 2013

SELECT origin, dest, a.name AS dest_name, flight, carrier
FROM flights AS f
JOIN airports AS a ON f.dest = a.faa
WHERE year = 2013 AND origin = 'ATW'
LIMIT 10;


```


Notice our use of the `AS` keyword in the `FROM` and `JOIN` clauses to create aliases (`f` for `flights` and `a` for `airports`). This is a critical best practice. It saves us from typing the full table name repeatedly, and more importantly, it prevents ambiguity errors. 

::: {.callout-note title="`dplyr` Equivalent of Above `SQL` Join"}

As mentioned above, every `SQL` join has a direct `dplyr` counterpart. Here, `SQL`'s `JOIN ... ON f.dest = a.faa` becomes `dplyr::inner_join(airports, by = c("dest" = "faa"))` - see the code below. 

Notice one important difference: in `SQL` we specify the join condition with `ON`, while in `dplyr` we specify it with the `by` argument, mapping the column name in the left table to the (possibly differently named) column in the right table. If we call `inner_join()` without a `by` argument, `dplyr` will attempt to guess the join column using any identically named columns in both tables — and will warn us about which column it chose. **Relying on this behavior is risky: always specify `by` explicitly**.

```{r}
#| eval: false

# equivalent query using dplyr's inner_join()
# please do not execute the following code, this will crash your machine

flights |>
  filter(year == 2013, origin == "ATW") |>
  inner_join(airports, by = c("dest" = "faa")) |>
  select(origin, dest, dest_name = name, flight, carrier)


```

:::



Data architectures often require us to chain **multiple joins** together to assemble a complete picture. What if we want the full names of the origin airport, the destination airport, and the carrier?

Because we need airport names for both the origin and the destination, we must join the `airports` table to our query twice. To do this, we must give the `airports` table two distinct aliases.


```{sql}
--| connection: con_air

-- access both airport and carrier names for flights that flew out of Appleton in 2013

SELECT origin, a2.name AS origin_name, 
       dest, a1.name AS dest_name, 
       flight, 
       f.carrier, c.name AS carrier_name
FROM flights AS f
JOIN airports AS a1 ON f.dest = a1.faa
JOIN airports AS a2 ON f.origin = a2.faa
JOIN carriers AS c ON f.carrier = c.carrier
WHERE origin = 'ATW' AND year = 2013
LIMIT 10;

```


### `LEFT JOIN`

While inner joins find intersections, `LEFT JOIN` is exceptionally useful for finding missing or mismatched data.

Let us intentionally look for destination airport codes in the `flights` table that do not have a corresponding entry in the `airports` table. We do this by performing a `LEFT JOIN` and filtering for rows where the airport name resolves to `NULL`.

```{sql}
--| connection: con_air

-- using LEFT JOIN

SELECT year, month, day, origin, dest,
      a.name AS dest_name,
      flight, carrier
FROM flights AS f
LEFT JOIN airports AS a ON f.dest = a.faa
WHERE a.name IS NULL
LIMIT 10;

```


In applied contexts, queries like this are fundamental to data auditing. In this specific dataset, this query reveals flights traveling to US territories like Puerto Rico (BQN, SJU) or the Virgin Islands (STT). Because they operate slightly outside the standard domestic FAA categorization used to build our `airports` table, they show up as missing values in our join.


::: {.callout-tip title="`dplyr::anti_join()` Pattern"}
The query above — a `LEFT JOIN` followed by a filter for `IS NULL` on the right-hand table — is common enough to have its own name: an **anti-join**. It answers the question "which rows in the first table have **no match at all** in the second table?" This is exactly what `dplyr::anti_join()` does in one step:

```{r}
#| eval: false

# equivalent query using dplyr's anti_join()
# please do not execute the following code, this will crash your machine

flights |>
 anti_join(airports, by = c("dest" = "faa")) |>
 select(year, month, day, origin, dest_name = name, flight, carrier)

```

Under the hood, `dplyr::anti_join()` is implemented using this same left-join-plus-filter logic. Recognizing the pattern is useful because anti-joins come up constantly in real data work: finding orders with no matching customer record, students with no matching enrollment record, or — as here — destination codes with no matching airport record.

:::

::: {.callout-note title="Handling Missing Values in `SQL`"}
Users of `R` are used to `NA` and functions like `is.na()`. `SQL` has a related but stricter concept: `NULL`, representing an **unknown or missing value**. The critical difference is that `NULL` cannot be compared using the standard `=` operator. A test like `a.name = NULL` does not evaluate to `TRUE` or `FALSE` — it evaluates to `UNKNOWN`, and rows where this happens are silently excluded from the result, with no error raised. To correctly test for missing values, `SQL` provides the dedicated keywords `IS NULL` and `IS NOT NULL`. 

Notice the difference in the `WHERE` clause between the queries above and below..

```{sql}
--| connection: con_air

SELECT year, month, day, origin, dest,
      a.name AS dest_name,
      flight, carrier
FROM flights AS f
LEFT JOIN airports AS a ON f.dest = a.faa
WHERE a.name = NULL
LIMIT 10;

```

:::


### `RIGHT JOIN`

`RIGHT JOIN` is simply the mirror image of `LEFT JOIN`: it keeps all rows from the table on the **right** side of the join, along with matching rows from the table on the left. In fact, any `RIGHT JOIN` can be rewritten as an equivalent `LEFT JOIN` just by swapping the order of the two tables — and vice versa. The results of the `LEFT JOIN` query above and the `RIGHT JOIN` query below are identical.

Because of this symmetry, most `SQL` style guides recommend sticking to `LEFT JOIN` for consistency and simply reordering tables as needed, rather than mixing `LEFT` and `RIGHT` in the same project. This is also why `dplyr` provides `right_join()` mainly for completeness — in practice, most `R` users express the same logic with `left_join()` and swap the table order instead.


```{sql}
--| connection: con_air

-- same result as our LEFT JOIN example above, but written as a RIGHT JOIN
-- with the tables swapped

SELECT year, month, day, origin, dest,
      a.name AS dest_name,
      flight, carrier
FROM airports AS a
RIGHT JOIN flights AS f ON f.dest = a.faa
WHERE a.name IS NULL
LIMIT 10;

```



## Unions (Vertical Integration)

While joins add columns to our result set, set operations add rows. `UNION` allows us to append the results of one query directly beneath the results of another, much like `dplyr::bind_rows()`.

Let us extract a dataset combining all flights that flew from ATW to MSP on April 6, 2013, with all flights that flew from JFK to ORD on the same day.

```{sql}
--| connection: con_air

-- extract all flights that flew from ATW to MSP and from JFK to ORD on April 6, 2013

(SELECT year, month, day, origin, dest, flight, carrier
FROM flights
WHERE year = 2013 AND month = 4 AND day = 6 AND origin = 'ATW' AND dest = 'MSP')

UNION

(SELECT year, month, day, origin, dest, flight, carrier
FROM flights
WHERE year = 2013 AND month = 4 AND day = 6 AND origin = 'JFK' AND dest = 'ORD')

LIMIT 10;


```


::: {.callout-warning title="Union Edge Cases"}
When using `UNION`, we must be aware of two strict database rules:

* **Column Matching**: Unlike `R`'s `bind_rows()`, which politely aligns columns by their names and fills in `NA`s where they don't match, `SQL` `UNION` does not check column names. It blindly stacks the tables based on the number of columns and their data types. If query A returns a string in column 1, and query B returns an integer in column 1, the query will fail.

* **Deduplication**: By default, `UNION` removes exact duplicate rows from the final result set. Searching for and removing duplicates is computationally expensive. If we know our datasets are mutually exclusive (as they are above) or if we want to retain duplicates, we should use `UNION ALL`, which simply appends the rows and executes much faster.

:::


`UNION` is only one of three standard set operations available in `SQL`. The other two — `INTERSECT` and `EXCEPT` — follow the exact same syntax rules (each side must be a complete `SELECT` statement with matching columns and types), but combine rows differently:

* **`INTERSECT`** keeps only the rows that appear in **both** result sets — analogous to `dplyr::intersect()`.

```{sql}
--| connection: con_air

-- destinations served from ORD in 2013 that were ALSO served in 2015
-- (INTERSECT keeps only rows that appear in BOTH result sets)

(SELECT DISTINCT dest FROM flights WHERE origin = 'ORD' AND year = 2013)

INTERSECT

(SELECT DISTINCT dest FROM flights WHERE origin = 'ORD' AND year = 2015);

```

* **`EXCEPT`** keeps rows from the first result set that do **not** appear in the second — analogous to `dplyr::setdiff()`.

```{sql}
--| connection: con_air

-- destinations served from ORD in 2013 that were NOT served in 2015
-- (EXCEPT keeps rows from the first result set that do NOT appear in the second)

(SELECT DISTINCT dest FROM flights WHERE origin = 'ORD' AND year = 2013)

EXCEPT

(SELECT DISTINCT dest FROM flights WHERE origin = 'ORD' AND year = 2015);

```


The two queries above ask complementary questions about Chicago O'Hare's route network: which destinations were served in both 2013 and 2015 (`INTERSECT`), and which destinations were served in 2013 but dropped by 2015 (`EXCEPT`). 

Together with `UNION`, these three operations round out our vertical-integration toolkit: 

* `UNION` stacks rows together, 
* `INTERSECT` finds their overlap, and 
* `EXCEPT` finds their difference.


## Subqueries

As our analytical questions grow more sophisticated, we often encounter situations where we cannot get our answer in a single pass. We need to evaluate a query, hold the result in memory, and then query that result.

We do this using subqueries. A **subquery** is simply a complete `SELECT` statement wrapped in parentheses and placed inside the `FROM` or `WHERE` clause of a parent query.

Let us return to the `UNION` query we just wrote above. What if we don't want to see the rows, but rather we just want a count of the total number of flights across both of those specific routes? We can wrap the entire `UNION` in parentheses and run a `COUNT(*)` over it.

```{sql}
--| connection: con_air

-- count the number of flights that satisfy the criteria above

SELECT COUNT(*) AS num_flights
FROM 
((SELECT year, month, day, origin, dest, flight, carrier
FROM flights
WHERE year = 2013 AND month = 4 AND day = 6 AND origin = 'ATW' AND dest = 'MSP')
UNION
(SELECT year, month, day, origin, dest, flight, carrier
FROM flights
WHERE year = 2013 AND month = 4 AND day = 6 AND origin = 'JFK' AND dest = 'ORD')
) AS temp_subquery;

```

::: {.callout-tip title="The Mandatory Alias"}
When we place a subquery inside a `FROM` clause, we are essentially creating a temporary table on the fly (often called a **derived table**). In `MySQL` and `MariaDB`, every derived table must have its own alias, even if we never explicitly refer to that alias in the `SELECT` statement. Notice the `AS temp_subquery` at the very end of our code block; without it, the `SQL` engine would throw a syntax error.
:::


The subquery in our previous example was used as a **derived table** — it stood in for an entire table in the `FROM` clause. Subqueries have a second, equally common use: appearing inside a `WHERE` clause to filter rows based on the result of another query.

In the following example, the **inner** query runs first, independently, and returns a list of carrier codes whose full name contains the word "Airlines" (using the `LIKE` operator together with the `%` wildcard, which matches any sequence of characters). The **outer** query then uses `IN` to keep only `flights` rows whose `carrier` appears in that list. 

```{sql}
--| connection: con_air

-- flights operated by any carrier whose name contains "Airlines"

SELECT year, month, day, origin, dest, carrier, flight
FROM flights
WHERE carrier IN (
  SELECT carrier
  FROM carriers
  WHERE name LIKE '%Airlines%'
)
LIMIT 10;

```

Conceptually, this is similar to first running `carriers |> filter(str_detect(name, "Airlines")) |> pull(carrier)` in `R`, and then using the result inside a `flights |> filter(carrier %in% ...)` call — except `SQL` performs both steps as a single query sent to the server.

::: {.callout-tip title="A Helpful Mental Model"}
A `FROM`-clause subquery builds a **temporary table**; a `WHERE`-clause subquery builds a **temporary list of values** to filter against.
:::

## which join/set operation/subquery to use?

In this chapter, we have been introduced to several distinct ways of combining data across tables and queries. As with the three `SQL`-in-`R` methods from Chapter 2, a short decision guide is more useful in practice than memorizing definitions:

* **Use an `INNER JOIN`** when we only want rows that have a match in both tables — for example, flights whose destination airport is known.

* **Use a `LEFT JOIN`** when we want to keep every row from your primary table regardless of whether a match exists — and especially when **auditing** for missing or unmatched data (our anti-join example above).

* **Use `UNION` / `UNION ALL`** when we are stacking comparable rows from two or more separately filtered queries into one result set.

* **Use `INTERSECT` or `EXCEPT`** when our question is about **overlap** or **difference** between two sets of values, rather than about combining full rows.

* **Use a subquery** when a question genuinely requires two passes over the data — first computing an intermediate result, then filtering or aggregating based on it — and a single `WHERE`/`GROUP BY`/`HAVING` combination cannot express the logic directly.

As our queries grow more complex, we will often combine several of these tools in a single analysis: joining tables to bring in descriptive names, filtering with a subquery, and aggregating the result — much as we did across the above sections.



## terminate connection

As always, we conclude our workflow by responsibly closing our database connections to free up memory on both our local machine and the remote server.

```{r}
# terminate the SQL connection

DBI::dbDisconnect(con_air, shutdown = TRUE)

```



