---
title: "Text Analysis II - Case Study of Donald Trump's tweets during the 2016 US presidential election"
format: pdf
editor: visual
---

# ------------------------------------------------------------------------------
# resources

* [Trump Twitter Archive](https://www.thetrumparchive.com/)

* [Todd Vaziri's Tweet](https://x.com/tvaziri/status/762005541388378112/photo/1)

* [David Robinson's Blog Post](http://varianceexplained.org/r/trump-tweets/)

* [Text Mining With R](https://www.tidytextmining.com/)


# ------------------------------------------------------------------------------
# packages

```{r}
library(tidyverse)
library(scales)
library(tidytext)
library(textdata)
library(dslabs)
```


# ------------------------------------------------------------------------------
# data

Our data come from the *dslabs* package. 

```{r}

data("trump_tweets")

?trump_tweets

glimpse(trump_tweets)

```


```{r}

range(trump_tweets$created_at)

```


```{r}

trump_tweets$text[16413]

```


```{r}

trump_tweets |> 
  count(source) |> 
  arrange(desc(n)) 

```


For our case study, we are interested in what happened during the 2016 campaign, so for this analysis we will focus on what was tweeted between the day Trump announced his campaign and election day. Also, we only keep tweets from Android or iPhone, and filter out retweets.

```{r}

campaign_tweets <- trump_tweets |> 
  filter(source %in% paste("Twitter for", c("Android", "iPhone")) &
           created_at >= ymd("2015-06-17") & 
           created_at < ymd("2016-11-08")) |>
  mutate(source = str_remove(source, pattern = "Twitter for ")) |>
  filter(!is_retweet) |>
  arrange(created_at) |> 
  as_tibble()

```


```{r}

campaign_tweets |>
  count(source)

```


# ------------------------------------------------------------------------------
# some initial exploration

We explore the possibility that two different groups were tweeting from these devices. 

```{r}

campaign_tweets |>
  mutate(hour = hour(with_tz(created_at, "EST"))) |>
  count(source, hour)

```


```{r}

android_vs_iphone_hourly_props <- campaign_tweets |>
  mutate(hour = hour(with_tz(created_at, "EST"))) |>
  count(source, hour) |>
  group_by(source) |>
  mutate(percent = n / sum(n)) 

android_vs_iphone_hourly_props

```


```{r}

android_vs_iphone_hourly_props |>
  ggplot(aes(x = hour, y = percent, color = source)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Hour of day (EST)", y = "% of tweets", color = "")

```


# ------------------------------------------------------------------------------
# tokens

```{r}

i <- 3008

campaign_tweets$text[i]

```


```{r}

campaign_tweets[i,] |> 
  tidytext::unnest_tokens(output = word, input = text, token = "words") |>
  dplyr::pull(word) 

```


```{r}

links_to_pics <- "(https://t.co/[A-Za-z\\d]+)|(http://t.co/[A-Za-z\\d]+)"

campaign_tweets[i,] |> 
  mutate(text = str_remove_all(text, pattern = links_to_pics)) |> 
  tidytext::unnest_tokens(output = word, input = text, token = "words") |>
  dplyr::pull(word) 

```


```{r}

tweet_words <- campaign_tweets |>
  mutate(text = str_remove_all(text, pattern = links_to_pics))  |>
  tidytext::unnest_tokens(output = word, input = text, token = "words")

```


```{r}

tweet_words |> 
  count(word) |>
  arrange(desc(n))

```


#### stop words

```{r}

data(stop_words)

?stop_words

```


```{r}

tweet_words <- campaign_tweets |> 
  mutate(text = str_remove_all(text, pattern = links_to_pics))  |>
  tidytext::unnest_tokens(output = word, input = text, token = "words") |> 
  filter(!(word %in% stop_words$word))

tweet_words |> 
  count(word) 

```


```{r}

tweet_words <- campaign_tweets |> 
  mutate(text = str_remove_all(text, links_to_pics))  |>
  tidytext::unnest_tokens(output = word, input = text, token = "words") |>
  filter(!(word %in% stop_words$word)) |> 
  filter(!(str_detect(word, pattern = "^\\d+")))   

tweet_words |> 
  count(word) 

# |>
#   arrange(desc(n))

```


#### further exploration

```{r}

android_vs_iphone <- tweet_words |>
  count(word, source) 

android_vs_iphone

```


```{r}

android_vs_iphone <- tweet_words |>
  count(word, source) |>
  pivot_wider(names_from = "source", values_from = "n", values_fill = 0) 

android_vs_iphone

```


```{r}

android_vs_iphone <- tweet_words |>
  count(word, source) |>
  pivot_wider(names_from = "source", values_from = "n", values_fill = 0) |>
  mutate(p_a = (Android + 0.5)/(sum(Android) + 0.5), 
         p_i = (iPhone + 0.5)/(sum(iPhone) + 0.5),  
         ratio = p_a / p_i)

android_vs_iphone

```


```{r}

# for words appearing at least 100 times in total, here are the ten highest percent differences for Android

top_10_android <- android_vs_iphone |> 
  filter(Android + iPhone >= 100) |> 
  arrange(desc(ratio)) |> 
  slice_head(n = 10)

top_10_android

```


```{r}

# and the top ten for iPhone

top_10_iphone <- android_vs_iphone |> 
  filter(Android + iPhone >= 100) |>  
  arrange(ratio)|> 
  slice_head(n = 10)

top_10_iphone

```


```{r}

top_10_android |> 
  rbind(top_10_iphone) |> 
  dplyr::select(word, iPhone, Android, ratio) |> 
  mutate(log_ratio = log(ratio)) |> 
  mutate(device = rep(c("Android", "iPhone"), each = 10)) |> 
  ggplot() +
  geom_col(aes(x = log_ratio, y = fct_reorder(word, log_ratio), fill = device))
  

```


# ------------------------------------------------------------------------------
# sentiment analysis

```{r}

tidytext::get_sentiments("bing")

tidytext::get_sentiments("afinn")

tidytext::get_sentiments("nrc")

```


```{r}

nrc <- tidytext::get_sentiments("nrc")

nrc |> count(sentiment)

```


```{r}

tweet_words |> 
  left_join(nrc, by = "word") |> 
  dplyr::select(source, word, sentiment) |> 
  sample_n(5)

```


```{r}

sentiment_counts <- tweet_words |>
  left_join(nrc, by = "word", relationship = "many-to-many") |>
  count(source, sentiment) 

sentiment_counts

```


```{r}

sentiment_counts <- tweet_words |>
  left_join(nrc, by = "word", relationship = "many-to-many") |>
  count(source, sentiment) |>
  pivot_wider(names_from = "source", values_from = "n") |>
  mutate(sentiment = replace_na(sentiment, replace = "none"))

sentiment_counts

```


```{r}

sentiment_counts <- sentiment_counts |>
  mutate(p_a = Android/sum(Android), 
         p_i = iPhone/sum(iPhone), 
         ratio = p_a/p_i) |> 
  arrange(desc(ratio))

sentiment_counts

```


```{r}

ordered_levels <- sentiment_counts$sentiment

android_vs_iphone |>
  filter(Android + iPhone >= 10) |>
  left_join(nrc, by = "word") |>
  mutate(sentiment = factor(sentiment, levels = sentiment_counts$sentiment)) |>
  group_by(sentiment) |>
  slice_max(abs(log(ratio)), n = 10) |>
  ungroup() |> 
  mutate(word = reorder(word, -ratio)) |>
  ggplot(aes(word, ratio, fill = ratio < 1)) +
  geom_col(show.legend = FALSE) +
  scale_y_log10() +
  facet_wrap(~sentiment, scales = "free_x", nrow = 2) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

```


# ------------------------------------------------------------------------------
