---
title: "Case Study of Shakepearse's Macbeth using Regular Expressions"
execute: 
  eval: false
  warning: false
  message: false
  cache: true
---



## resources

- [Text Mining with R](https://www.tidytextmining.com/)

- [Modern Data Science with R (3e) - Chapter 19](https://mdsr-book.github.io/mdsr3e/)



## packages

```{r}
library(tidyverse)
```



## data

Our data comes from [Project Gutenberg](https://www.gutenberg.org/), a large repository of free eBooks. We will be working with the text of Shakespeare’s play Macbeth, which is available at https://www.gutenberg.org/cache/epub/1129/pg1129.txt. The text is in the public domain, so we are free to use it for our analysis.

```{r}
library(gutenbergr)

data("gutenberg_metadata")

macbeth_all <- gutenberg_metadata |>
  filter(str_detect(title, "Macbeth"))

book <- gutenberg_download(1129, meta_fields = "title")

```

For our case study, we will be working with the raw text of the play, which is available in the `mdsr` package.

```{r}

library(mdsr)

data(Macbeth_raw)

?Macbeth_raw

```

```{r}

macbeth <- Macbeth_raw |>
  str_split("\r\n") |>
  pluck(1)

length(macbeth)

macbeth[300:310]

```



## strings and regular expressions

```{r}
# how many times does the character Macbeth speak in the play?

macbeth_lines <- macbeth |>
  str_subset("  MACBETH")

length(macbeth_lines)

head(macbeth_lines)

```

```{r}

macbeth |>
  str_which("  MACBETH")

```

## about regular expressions grammar

### metacharacter

```{r}

macbeth |>
  str_subset("MAC.") |>
  head()  

```

```{r}

macbeth |>
  str_subset("MACBETH\\.") |>
  head()

```

### character sets

```{r}

macbeth |>
  str_subset("MAC[B-Z]") |>
  head()

```

### alternation

```{r}

macbeth |>
  str_subset("MAC(B|D)") |>
  head()

```

### anchors (power-dollar)

```{r}

macbeth |>
  str_subset("^  MAC[B-Z]") |>
  head()

```

```{r}

macbeth |>
  str_subset("\\.$") |>
  head()

```

### repetitions

```{r}

# ?  

macbeth |>
  str_subset("^ ?MAC[B-Z]") |>
  head()

```

```{r}

# * 

macbeth |>
  str_subset("^ *MAC[B-Z]") |>
  head()

```

```{r}

# + 

macbeth |>
  str_subset("^ +MAC[B-Z]") |>
  head()

```



## a case study - life and death in Macbeth and speaking patterns

```{r}

macbeth_chars <- tribble(
  ~name, ~regexp,
  "Macbeth", "  MACBETH\\.",
  "Lady Macbeth", "  LADY MACBETH\\.",
  "Banquo", "  BANQUO\\.",
  "Duncan", "  DUNCAN\\.",
) |>
  mutate(speaks = purrr::map(regexp, str_detect, string = macbeth))

```

```{r}

speaker_freq <- macbeth_chars |>
  unnest(cols = speaks) |>
  mutate(
    line = rep(1:length(macbeth), 4),
    speaks = as.numeric(speaks)
  ) |>
  filter(line > 218 & line < 3172)

glimpse(speaker_freq)

```

```{r}

acts <- tibble(
  line = str_which(macbeth, "^ACT [I|V]+"), 
  line_text = str_subset(macbeth, "^ACT [I|V]+"),
  labels = str_extract(line_text, "^ACT [I|V]+")
)

```

```{r}

ggplot(data = speaker_freq, aes(x = line, y = speaks)) + 
  geom_smooth(
    aes(color = name), method = "loess", 
    se = FALSE, span = 0.4
  ) + 
  geom_vline(
    data = acts, 
    aes(xintercept = line), 
    color = "darkgray", lty = 3
  ) + 
  geom_text(
    data = acts, 
    aes(y = 0.085, label = labels), 
    hjust = "left", color = "darkgray"
  ) + 
  ylim(c(0, NA)) + 
  xlab("Line Number") +
  ylab("Proportion of Speeches") + 
  scale_color_colorblind()

```


